---
title: "FinalRuleEDA"
author: "Chenery Lowe"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=TRUE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)

# set working directory
setwd("C:/Users/clowe7/OneDrive - Stanford/Final Rule Analysis") #***You'll want to change the working directory to reflect the location where your files are stored

# install packages
library(dplyr)
library(readxl)
library(tidyverse)
library(ggplot2)
library(tidyverse)
library(arsenal)

# load data file ***Make sure that the file is in the working directory that you set above; otherwise, you will need to specify the full pathway
data<-read.csv("102 Survey_AssessingGeneticCoun-ForDataAnalysis_DATA_LABELS_2024-02-06_1407 (1).csv")

data_raw<-read.csv("AssessingGeneticCoun-ForDataAnalysis_DATA_2024-02-25_1458.csv")

# ***May need to clean, convert, or recode variables - here's one example

# convert years experience as ordered factor variable
# Create a vector of levels in correct order
levels <- c("1- 4 years", "5- 9 years", "10- 14 years", "15- 19 years", "20- 24 years", "25- 29 years")

# Convert to an ordered factor
data$How.many.years.experience.do.you.have.as.a.practicing.genetic.counselor..round.to.the.nearest.year.. <- factor(data$How.many.years.experience.do.you.have.as.a.practicing.genetic.counselor..round.to.the.nearest.year.., levels = levels, ordered = TRUE)



# set up labels for the participant characteristics table ***This list isn't complete - add the rest of the variable names here
labs<- c(What.is.your.age.in.years. = "Age")

```

# Summary table
```{r, results='asis'}
# select the relevant variables
desc<-data%>%
  select(What.is.your.age.in.years.) # ***add more demographic variables

# variable labels
labels(desc)<-labs

#create formula
myvars<-names(desc)
tmp<-formulize(X=myvars[1:(ncol(desc))])

# *** if you wanted to stratify by a variable, you could do this: tmp<-formulize("***variable name", myvars[1:(ncol(desc))])

summary(tableby(tmp, data = desc,
              digits = 2L))


```
# recode variables for workplace policies variable - analysis 1a (presence vs. absence of the policy)
```{r}
#Analysis 1a: The existence of direct contact, institutional policies, 
#dept/division/workgroup policies, blocking policies, and changes to pt portal 
#policies are associated with less perceived disruption to workflow, less 
#perceived pt harms, and more perceived pt benefits.

# Recode policies as 0 if it does not exist or not sure, NA if NA, and 1 if anything else
policies_dept_1a <- ifelse(is.element(data$policies_dept, notexist),0,
                           (ifelse(is.element(data$policies_dept,na),NA,1)))

policies_institutional_1a <- ifelse(is.element(data$policies_instituitional, notexist),0,
                                    (ifelse(is.element(data$policies_instituitional,na),NA,1)))

policies_portal_1a <- ifelse(is.element(data$policies_portal, notexist),0,
                             (ifelse(is.element(data$policies_portal,na),NA,1)))

```


# recode variables for analysis 1b (GC involved in policy vs. not involved vs. no known policy)
```{r}
#Analysis1b: The involvement of GCs in these policies is associated with less 
#perceived disruption to workflow, less perceived pt harms, and more perceived 
#pt benefits.

# recode variables for analysis 1b (GC involved in policy vs. not involved vs. no known policy)
policies_dept_1b <- ifelse(is.element(data$policies_dept, notexist),0,
                           (ifelse(is.element(data$policies_dept,na),NA,
                                   (ifelse(is.element(data$policies_dept,GCnot),1,2)))))

policies_institutional_1b <- ifelse(is.element(data$policies_instituitional, notexist),0,
                                    (ifelse(is.element(data$policies_instituitional,na),NA,
                                            (ifelse(is.element(data$policies_instituitional,GCnot),1,2)))))

policies_portal_1b <- ifelse(is.element(data$policies_portal, notexist),0,
                             (ifelse(is.element(data$policies_portal,na),NA,
                                     (ifelse(is.element(data$policies_portal,GCnot),1,2)))))
```


# recode variables for outcomes (harms, benefits, and workflow impact)
```{r}
#perceived harms
pt_reaction_RC <- recode(data$pt_reaction,"5" = 5, "4" = 4, "3" = 3, "2" = 2, 
                         "1" = 1,"6" = NA_real_)
pt_understand_RC <- recode(data$pt_understand,"5" = 5, "4" = 4, "3" = 3, "2" = 2, 
                         "1" = 1,"6" = NA_real_)
pt_contact_RC <- recode(data$pt_contact,"5" = 5, "4" = 4, "3" = 3, "2" = 2, 
                           "1" = 1,"6" = NA_real_)

df_harms <- data.frame(pt_reaction_RC,pt_understand_RC,pt_contact_RC)
sums_harms <- (rowSums(df_harms))/3
summary(sums_harms)


#perceived benefits
pt_empower_RC <- recode(data$pt_empower,"5" = 5, "4" = 4, "3" = 3, "2" = 2, 
                           "1" = 1,"6" = NA_real_)
pt_processing_RC <- recode(data$pt_processing,"5" = 5, "4" = 4, "3" = 3, "2" = 2, 
                           "1" = 1,"6" = NA_real_)
pt_assurance_RC <- recode(data$pt_assurance,"5" = 5, "4" = 4, "3" = 3, "2" = 2, 
                           "1" = 1,"6" = NA_real_)
pt_action_RC <- recode(data$pt_action,"5" = 5, "4" = 4, "3" = 3, "2" = 2, 
                           "1" = 1,"6" = NA_real_)
pt_share_RC <- recode(data$pt_share,"5" = 5, "4" = 4, "3" = 3, "2" = 2, 
                           "1" = 1,"6" = NA_real_)

df_benefits <- data.frame(pt_empower_RC,pt_processing_RC,pt_assurance_RC,pt_action_RC,pt_share_RC)
sums_benefits <- (rowSums(df_benefits))/5
summary(sums_benefits)

#workplace effects
checking_RC <- recode(data$checking,"5" = 5, "4" = 4, "3" = 3, "2" = 2, 
                        "1" = 1,"6" = NA_real_)
timing_RC <- recode(data$timing,"5" = 5, "4" = 4, "3" = 3, "2" = 2, 
                      "1" = 1,"6" = NA_real_)
communications_RC <- recode(data$communications,"5" = 5, "4" = 4, "3" = 3, "2" = 2, 
                      "1" = 1,"6" = NA_real_)
change_RC <- recode(data$change,"5" = 5, "4" = 4, "3" = 3, "2" = 2, 
                      "1" = 1,"6" = NA_real_)
coordination_dept_RC <- recode(data$coordination_dept,"5" = 5, "4" = 4, "3" = 3, "2" = 2, 
                    "1" = 1,"6" = NA_real_)
coordination_other_RC <- recode(data$coordination_other,"5" = 5, "4" = 4, "3" = 3, "2" = 2, 
                    "1" = 1,"6" = NA_real_)
sending_results_RC <- recode(data$sending_results,"5" = 1, "4" = 2, "3" = 3, "2" = 4, 
                    "1" = 5,"6" = NA_real_)

```





# explore the distribution for numerical variables (workflow variables, in this example)
```{r}
df_workplace <- data.frame(checking_RC,timing_RC,communications_RC,change_RC,coordination_dept_RC, coordination_other_RC,sending_results_RC)
sums_workplace <- (rowSums(df_workplace)/7)
summary(sums_workplace)

#sums
summary(sums_benefits)
summary(sums_harms)
summary(sums_workplace)

hist(sums_benefits)
hist(sums_harms)
hist(sums_workplace)

# create multiple histogram
hist(sums_harms, col='red', xlim=c(1, 5))
hist(sums_benefits, col='green', add=TRUE)
hist(sums_workplace, col='blue', add=TRUE)

```

# To-do next:

## Factor analysis for outcome variables

### Perceived benefits

```{r}
library(psych)
library(corrplot)
library(car)

# subset perceived benefits variables
ben<-data_raw%>%subset(select = c(pt_empower:pt_share)) %>% mutate_all(as.numeric) %>% mutate_all(na.omit)

# correlation matrix
datamatrix<-cor(ben)
corrplot::corrplot(datamatrix, method = "number")

# analyze factorability of the data
X<-ben

#Kaiser-Meyer Olkin
KMO(r=cor(X))

#Bartlett's test of sphericity
cortest.bartlett(X)
det(cor(X))

# scree plot
library(ggplot2)
fafitfree <- fa(ben,nfactors = ncol(X), rotate = "none")
n_factors <- length(fafitfree$e.values)
scree     <- data.frame(
  Factor_n =  as.factor(1:n_factors), 
  Eigenvalue = fafitfree$e.values)
ggplot(scree, aes(x = Factor_n, y = Eigenvalue, group = 1)) + 
  geom_point() + geom_line() +
  xlab("Number of factors") +
  ylab("Initial eigenvalue") +
  labs( title = "Scree Plot", 
        subtitle = "(Based on the unreduced correlation matrix)")

parallel<-fa.parallel(X)

# 1 factor
factanal_1 <- factanal(X, factors=1, scores = c("regression"), rotation = "varimax")
print(factanal_1)

# 2 factors
factanal_2 <- factanal(X, factors=2, scores = c("regression"), rotation = "varimax")
print(factanal_2)

# check internal consistency of benefits variables (Cronbach's alpha)

# correlation matrix
datamatrix<-cor(ben)
corrplot::corrplot(datamatrix, method = "number")

# Cronbach's alpha, allowing negatively correlated items to be reversed
psych::alpha(X, check.keys = TRUE)

# create benefits score
data_raw$benefits <- rowSums(data_raw[,c("pt_empower", "pt_processing", "pt_assurance", "pt_action", "pt_share")], na.rm = TRUE)


```
### Do the same for perceived harms and perceived impact on workflow



## Bivariate analyses for presence/absence of policy (how is each individual predictor related to each individual outcome and outcome categories) – logistic and linear
```{r, results='asis'}
# to do a summary table for departmental/workgroup policies

# select the relevant variables
desc<-data_raw%>%
  select(policies_dept_1a, pt_empower) # ***add more outcome variables

#convert numeric to factor variables, if needed ***may not be needed if converted to factor in an earlier step
desc$pt_empower<-as.factor(desc$pt_empower)

# variable labels
labels(desc)<-labs

#create formula
myvars<-names(desc)
tmp<-formulize("policies_dept_1a", myvars[2:(ncol(desc))])

summary(tableby(tmp, data = desc,
              digits = 2L))


```


## Bivariate analyses for GC involvement in policy – logistic and linear (similar to previous chunk)



## Multiple linear regression models relating predictors (resources, policies, changes to practice) to outcomes (impact on workflow, pt harms, pt benefits), adjusting for GC and workplace characteristics

```{r}
#subset the variables that you are considering using in the model ***change these as needed - maybe add workplace setting, years of training, etc.
tmp<- subset(data_raw, select = c("benefits", "policies_dept_1a", "ordering_provider", "position"))
             
#subset cases with non-missing data
tmp<-tmp[complete.cases(tmp),]

#define intercept-only model
intercept<-lm(benefits ~ 1, data = tmp)

#define model with all predictors
all<-lm(benefits~., data = tmp)

#perform forward and back stepwise regression
both<-step(intercept, direction= "both", scope = formula(all), trace = 0)

#view covariates selected by stepwise selection
both$coefficients

#fit model using these variables ***change these as needed
ben1a<-lm(benefits~ policies_dept_1a + ordering_provider + position, data = tmp)

#check for multicollinearity - calculate variance-inflation factor
car::vif(ben1a)

#view summary stats for selected model
summary(ben1a)

#assess normality
#regression diagnostics
library(ggfortify)
autoplot(ben1a)
shapiro.test(ben1a$residuals) #assess normality of residuals
```


